HyperLogLog: Exponentially Bad in Adversarial Settings

### 2022-02-06 docs/HLL_exp_bad/main.tex
clean up to results section and Redis more generally

We assume that at least $2^{t-1}\cdot m\ln m$ items have already been honestly inserted into the target HLL, for some integer $t$ with $1 \leq t \leq \ell$ which will be a parameter of our attack. We assume the S2 adversary somehow knows that this condition holds (e.g.\ through knowledge of when the HLL was initialised and the average rate at which items are inserted into it). Notice that an S3 adversary can use its ability to get cardinality estimates to obtain this knowledge; this is one situation where an S3 adversary can make good use of the extra information available to it as compared to an S2 adversary. In S4, the adversary can use its access to a snapshot to mount a more powerful attack; see Section~\ref{sec:nonemptyHLLS4}.

Under the condition that at least $2^{t-1}\cdot m\ln m$ items have already been honestly inserted, it is highly likely that every one of the $m=2^n$ buckets has counter value $M[j]$ with $M[j] \ge t$. This follows from an extension of the standard analysis of the coupon collector problem~\cite[Section 3.6]{MR95} under our usual assumption that $h$ behaves as a random function. To see this, note that when an item $x$ is inserted into a specific bucket $j$, the condition $M[j] \ge t$ holds with probability $2^{-t+1}$; we then need about $m\ln m$ such events to occur to ensure the desired condition holds in every bucket.

With $2^{t-1}\cdot m \ln m$ honest insertions, it is also highly likely that the small range correction (Hit Counting algorithm) will not be used, and the target HLL's cardinality estimate will be computed using the raw estimate.
%\[
%E = \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}.
%\]

@@

The S2 adversary knows $h$ so it selects items $x$ from ${\mathcal{X}}$ for which the $\ell$ rightmost bits of $h(x)$ start with up to $t-1$ ``0'' bits. In this case, since the relevant bucket $j$ already has $M[j] \ge t$, the bucket's counter is not updated, and the cardinality estimate of the HLL sketch remains unchanged.

@@

In scenarios S2 and S4 with an initially non-empty HLL sketch, we assume as for our earlier analysis of the standard HLL algorithm that all buckets have already been hit (this will be the case with high probability as soon as $m \log m$ items have been honestly inserted). Hence $C_0 = 0$ and $\sigma$ is evaluated at 0 in (\ref{eqn:ertl}), where its value is 0. This means that the estimate $\hat{\lambda}$ is computed using a denominator of the form:
\[
\sum_{k=1}^{\ell}C_k 2^{-k} + m \tau(1-C_{\ell+1}/m)2^{-\ell}.
\]
Now we can proceed exactly as in Sections~\ref{sec:nonemptyHLLS2} and~~\ref{sec:nonemptyHLLS4}, inserting items $x$ whose hashes have rightmost $\ell$ bits beginning with a ``1'' bit (and with a more relaxed condition in S4). These insertions do not alter the values of the $C_k$, and hence $\hat{\lambda}$ is unaffected. In S2, we expect to insert half the items in ${\mathcal{X}}$ without increasing the cardinality estimate $\hat{\lambda}$ at all. In S4, the analysis is the same as in Section~\ref{sec:nonemptyHLLS4}.

@@

We first focus on a classic HLL implementation as per the original HLL paper~\cite{hll}, namely the implementation of~\cite{clahll}.
We choose $h$ to be the 32-bit MurmurHash3 function from~\cite{murmur3code}; again the attacks do not depend on the specifics of what fixed hash function is used.
%(note that the library allows the hash function to be specified; we pick Murmur3 for our attack and consider it reasonable that the adversary should know the hash function in use).
We create an HLL sketch using $n=8$ and hence $m=256$ buckets and, for attacks in the setting of an initially non-empty target HLL sketch, initialise it with 1,000 random items representing honest users' data.

We then challenge the adversary to pick the largest possible number of items to add to the sketch from our set of  $2^{20}$ possible inputs while trying to keep the cardinality estimate as low as possible, in each of the scenarios S1, S2 and S4.

As a second experiment, and in order to get results that can be compared to the work of~\cite{hllvuln}, we mount our attack in their more restrictive setting: instead of picking from ${\mathcal{X}}$, the adversary has to choose what to insert from a subset of 250,000 random items that are given to it.

Finally, for completeness, we also run the attacks against HLL sketches filled with adversarial inputs only, thus that are initially empty. We vary the value of $B$ in the attack for scenario S1 to demonstrate some of the available trade-offs. For simplicity and to keep the attack as striking as possible, we set $B$ to 1 in the attacks on an initially empty HLL sketch in scenarios S2 and S4. All attacks were run 50 times and we report average outcomes. Our results are reported in Tables~\ref{table:tab1} and~\ref{table:tab2}, which detail the estimated cardinality at the beginning of the attack, the number of items added by the adversary, and the cardinality estimate after adding the adversary's items.

These empirical results confirm our analysis on several points.

First, when attacking a sketch already containing 1000 random items, we can see that the adversary is on average able to insert about half of the $2^{20}$ items in our starting set in scenario S2, and about 83\% in scenario S4. The cardinality estimate is, on average, increased slightly in S2 (from 1,001 to 1,025) and remains at the same starting value (1,007 on average) in S4. In both cases, we inserted more than half a million items whilst increasing the cardinality estimate by about $0.2\%$!  These results align well with our theoretical analysis.  
%In scenario S1 we run the attack designed for an initially empty target HLL sketch. It performs well in practice, for example, increasing the HLL cardinality estimate from 100 to 126????
%( $205$'$000 \approx 250'000 \times 0.82 \approx 250$'$000\times(1-\frac{1}{2\times H_c})$, with $H_c\approx\frac{1'000}{256}=3.9$)
%Here $0.83 \approx (1-\frac{1}{2\cdot H_c})$ with $H_c\approx\frac{1000}{256}=3.9$, so the results are also consistent with our theoretical analysis.
%% That analysis was now updated -- does it still match?
Second, when attacking an empty HLL sketch, we can see that in all scenarios, the final cardinality estimate is always exactly a rounded version of the Hit Counting estimator $m \cdot \ln (m/(m-B))$ after inserting a fraction $\frac{B}{m}$ of the items, as predicted by our analysis. For example, in scenario S1 with $B=128$, the final cardinality estimate is 177 (in agreement with the formula) while with $B=1$ in scenarios S2, S4, the final cardinality estimate is just 1 (again, as predicted by the formula).

We can also compare our results from Table~\ref{table:tab2} with results from the attack presented in~\cite[Section 5.1]{hllvuln}. In that paper, the authors target the Redis implementation of HLL and perform the attack against an empty sketch filled with adversarial inputs only. Given a set of 250,000 items, the adversary could choose ``74,390 distinct items and obtain an HLL estimate of only 15,780''~\cite{hllvuln}, achieving a five-fold reduction from the true cardinality after 4 iterations (resets). As explained in Section~\ref{sec:redis}, the Redis implementation analysed in~\cite{hllvuln} does not conform precisely to the original HLL algorithm as presented in Figure~\ref{fig:hll}. Also as discussed in Section~\ref{sec:redis}, the Redis implementation has changed its HLL algorithm several times in recent releases, but the authors of~\cite{hllvuln} do not specify which version was used in their experiments. This makes it difficult to make a precise comparison with the results of~\cite{hllvuln}. Nevertheless, the S1 adversary following our strategy against the classical HLL algorithm is able to add 977 and 125,091 distinct items while holding the cardinality estimate at 1 and 177, respectively. Compared to~\cite{hllvuln}, we can reduce the cardinality estimate by a factor of almost 1,000 instead of 5. This comes at the cost of performing more resets compared to~\cite{hllvuln}. We recall that in practice each reset may cause a delay (for example, it may involve rebooting a network device); we also recall that we consider S1 to be a somewhat artificial attack scenario.

@@

\begin{figure*}[htp]
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \caption{Attack results as function of t in scenario S2 with initially non-empty HLL sketch.}
        \includegraphics[width=0.9\textwidth]{figures/graph_S2.pdf}
    \end{minipage}\hfill
    \begin{minipage}{0.45\textwidth}
        \centering
        \caption{Attack results as function of t in scenario S4 with initially non-empty HLL sketch.}
        \includegraphics[width=0.9\textwidth]{figures/graph_S4.pdf} % second figure itself
     \end{minipage}
    \label{fig:graphs}
\end{figure*}

@@

\subsection{Results for Redis}
We also implemented our attacks for the HLL implementation used in Redis as described in Section~\ref{sec:redis}.
The hash function $h$ is fixed as the 64-bit MurmurHash3 function. The seed used internally is hardcoded and can be extracted by inspection of the openly-available Redis codebase.
Furthermore, the number of buckets is constant and set to 16,384. Knowledge of these parameters places us in the setting of an S2 adversary.
Since a classic Redis application separates the server holding the sketch from the client interacting with the sketch, we consider that attack scenarios S3 or S4, while possible, are not typical for Redis. 
Thus, we only consider scenario S2 for Redis.
For simplicity, we set attack parameter $\epsilon$ to $\frac{1}{2}$ in the attack on an initially empty HLL sketch.
The results are averaged over 50 iterations and reported in Table~\ref{table:tab3}.
Similarly to the case for classic HLL, our empirical results match with our theoretical analysis.

\begin{table*}[tb!]
\begin{center}
\caption{Attack results for HLL in Redis, averaged over 50 iterations.}
\begin{tabular}{| m{11em} | m{4em} | m{4.2em} |}
    \hline
    \textbf{Scenario} &  \multicolumn{2}{c|}{S2} \\ \hline
    %\textbf{HLL initially empty?} & no & yes \\ \hline
    \textbf{Original Card. Est.} & 0 & 139,967 \\ \hline
     \textbf{$\epsilon$} &  0.5 & - \\ \hline
    \textbf{\# Items added} & 524,210 & 524,454\\ \hline
    \textbf{Final Card. Est.} & 13,185 & 139,967 \\ \hline
\end{tabular}
\end{center}
\label{table:tab3}
\end{table*}
%\todo[inline]{Results from Redis or GBQ if we get them.}

@@

The authors thank Dennis Hofheinz and Anupama Unnikrishnan for useful discussions on simulation-based definitions, Tom Shrimpton for suggesting to look at GoogleBigQuery, and Eylon Yogev for drawing our attention to related work on streaming algorithms with adversarial inputs.
\fi

>>>>

We assume that at least $2^{t-1} m\ln m$ items have already been honestly inserted into the target HLL, for some integer $t$ with $1 \leq t \leq \ell$ which will be a parameter of our attack. We assume the S2 adversary somehow knows that this condition holds (e.g.\ through knowledge of when the HLL was initialised and the average rate at which items are inserted into it). Notice that an S3 adversary can use its ability to get cardinality estimates to obtain this knowledge; this is one situation where an S3 adversary can make good use of the extra information available to it as compared to an S2 adversary. In S4, the adversary can use its access to a snapshot to mount a more powerful attack; see Section~\ref{sec:nonemptyHLLS4}.

Under the condition that at least $2^{t-1} m\ln m$ items have already been honestly inserted, it is highly likely that every one of the $m=2^n$ buckets has counter value $M[j]$ with $M[j] \ge t$. This follows from the standard analysis of the coupon collector problem~\cite[Section 3.6]{MR95} under our usual assumption that $h$ behaves as a random function. To see this, note that when an item $x$ is inserted into a specific bucket $j$, the condition $M[j] \ge t$ holds with probability $q:=1/2^{t-1}$; so with a total of $N$ attempted insertions, we expect $Nq$ usable insertions, with a sharp concentration around this number. We need about $m\ln m$ usable insertions to occur to ensure we get one in every bucket; setting $Nq = m\ln m$ completes the argument.

With $2^{t-1}m \ln m$ honest insertions, it is also highly likely that the small range correction (Hit Counting algorithm) will not be used, and the target HLL's cardinality estimate will be computed using the raw estimate.
%\[
%E = \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}.
%\]

@@

The S2 adversary knows $h$ so it selects items $x$ from ${\mathcal{X}}$ for which the $\ell$ rightmost bits of $h(x)$ start with up to $t-1$ ``0'' bits. In this case, since the relevant bucket $j$ already has $M[j] \ge t$ (with high probability), the bucket's counter is not updated, and the cardinality estimate of the HLL sketch remains unchanged.

@@

In scenarios S2 and S4 with an initially non-empty HLL sketch, we assume that $2^{t-1}m \ln m$ items have been honestly inserted, as per our analysis in Section~\ref{sec:nonempty}. This ensures that, with high probability, we have $C_0 = C_1 = \cdots = C_{t-1} = 0$. In particular, since $C_0 = 0$, $\sigma$ is evaluated at 0 in (\ref{eqn:ertl}), where its value is 0. This means that the estimate $\hat{\lambda}$ is computed using a denominator of the form:
\[
\sum_{k=t}^{\ell}C_k 2^{-k} + m \tau(1-C_{\ell+1}/m)2^{-\ell}.
\]
Now we can proceed exactly as in Section~\ref{sec:nonempty}, inserting items $x$ whose hashes have rightmost $\ell$ bits beginning with up to $t-1$ ``0'' bits (and with a more relaxed condition in S4). With high probability, these insertions do not alter the values of the $C_k$, and hence $\hat{\lambda}$ should be unaffected. In S2, we expect to insert a fraction $1-1/2^t$ of the items in ${\mathcal{X}}$ without increasing the cardinality estimate $\hat{\lambda}$ at all. In S4, the analysis is the same as in Section~\ref{sec:nonemptyHLLS4}.

@@

We first focus on a classic HLL implementation as per the original HLL paper~\cite{hll}, namely the implementation of~\cite{clahll}.
We choose $h$ to be the 32-bit MurmurHash3 function from~\cite{murmur3code}; again the attacks do not depend on the specifics of what fixed hash function is used.
%(note that the library allows the hash function to be specified; we pick Murmur3 for our attack and consider it reasonable that the adversary should know the hash function in use).
We create an HLL sketch using $n=8$ and hence $m=256$ buckets. We run attacks in each of the scenarios S1, S2 and S4, with initially empty and non-empty HLL sketches. In the latter case, we initialise the sketch with $2^{t-1} m \ln m$ random items, where $t$ is an integer parameter that we vary in our experiments (cf.\ Section~\ref{sec:nonempty}). We then challenge the adversary to pick the largest possible number of items to add to the sketch from our set of  $2^{20}$ possible inputs while trying to keep the cardinality estimate as low as possible. As a second experiment, and in order to get results that can be compared to the work of~\cite{hllvuln}, we mount our attack in their chosen setting for scenario S1: instead of picking from ${\mathcal{X}}$, the adversary has to choose what to insert from a subset of 250,000 random items that are given to it. All attacks were run 50 times and we report average outcomes.

Our results are reported in Tables~\ref{table:tab1},~\ref{table:tab2} and Figures~\ref{fig:graphS2},~\ref{fig:graphS4}. The tables detail our results for the case of initially empty HLL sketches while the figures show results in S2 and S4 for initially non-empty HLL sketches.

%Finally, for completeness, we also run the attacks against HLL sketches filled with adversarial inputs only, thus that are initially empty. We vary the value of $B$ in the attack for scenario S1 to demonstrate some of the available trade-offs. For simplicity and to keep the attack as striking as possible, we set $B$ to 1 in the attacks on an initially empty HLL sketch in scenarios S2 and S4. All attacks were run 50 times and we report average outcomes. Our results are reported in Tables~\ref{table:tab1} and~\ref{table:tab2}. These tables detail the estimated cardinality at the beginning of the attack, the number of items added by the adversary, and the cardinality estimate after adding the adversary's items.

These experimental results broadly confirm our analyses. First, when attacking an empty HLL sketch (cf.\ Table~\ref{table:tab1}), we can see that in all scenarios, the final cardinality estimate is always exactly a rounded version of the Hit Counting estimator $m \cdot \ln (m/(m-B))$ after inserting a fraction $\frac{B}{m}$ of the items, as predicted by our analysis. For example, in scenario S1 with $B=128$, the final cardinality estimate is 177 (in agreement with the formula) while with $B=1$ in scenarios S2, S4, the final cardinality estimate is just 1 (again, as predicted by the formula). Moreover, the number of items we can insert on average aligns well with our theoretical predictions (e.g.\ $4106$ in S2 and S4, compared with a prediction of $2^{20}/2^8 = 4096$).

Second, Figures~\ref{fig:graphS2},~\ref{fig:graphS4} show very close agreement between the predictions made by our analysis in Section~\ref{sec:nonempty} and our experimental results for the situation where the HLL sketch is initially non-empty. The only significant difference is in scenario S2, where there is a small increase in the cardinality estimate, around 1-2\% on average, as a result of the attack, whereas the theoretical analysis suggests there should be no increase. The reason for this is that the insertion of the initial random $2^{t-1} m\ln m$ items is not guaranteed to produce a situation in which every bucket contains a value $t$ or greater -- this condition holds only with high probability.

We can also compare with the attack presented in~\cite[Section 5.1]{hllvuln}. In that paper, the authors target the Redis implementation of HLL and perform the attack against an empty sketch filled with adversarial inputs only. Given a set of 250,000 items, the adversary could choose ``74,390 distinct items and obtain an HLL estimate of only 15,780''~\cite{hllvuln}, achieving a five-fold reduction from the true cardinality after 4 iterations (resets). As explained in Section~\ref{sec:redis}, the Redis implementation analysed in~\cite{hllvuln} does not conform precisely to the original HLL algorithm as presented in Figure~\ref{fig:hll}. Also as discussed in Section~\ref{sec:redis}, the Redis implementation has changed its HLL algorithm several times in recent releases, but the authors of~\cite{hllvuln} do not specify which version was used in their experiments. This makes it difficult to make a precise comparison with the results of~\cite{hllvuln}. Nevertheless, the S1 adversary following our strategy against the classical HLL algorithm is able to add 977 and 125,091 distinct items while holding the cardinality estimate at 1 and 177, respectively (see Table~\ref{table:tab2}). Compared to~\cite{hllvuln}, we can reduce the cardinality estimate by a factor of almost 1,000 instead of 5. This comes at the cost of performing significantly more resets compared to~\cite{hllvuln} (more than 1,000 compared to 3). We recall that in practice each reset may cause a delay (for example, it may involve rebooting a network device); we also recall that we consider S1 to be a somewhat artificial attack scenario.

@@

\begin{figure*}[htp]
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \caption{Attack results as function of t in scenario S2 with initially non-empty HLL sketch.}\label{fig:graphS2}
        \includegraphics[width=0.9\textwidth]{figures/graph_S2.pdf}
    \end{minipage}\hfill
    \begin{minipage}{0.45\textwidth}
        \centering
        \caption{Attack results as function of t in scenario S4 with initially non-empty HLL sketch.}\label{fig:graphS4}
        \includegraphics[width=0.9\textwidth]{figures/graph_S4.pdf} % second figure itself
     \end{minipage}
\end{figure*}

@@

\subsection{Results for Redis}
We also implemented our attacks for the HLL implementation used in Redis as described in Section~\ref{sec:redis}.
The hash function $h$ is fixed as the 64-bit MurmurHash3 function. The seed used internally is hardcoded and can be extracted by inspection of the openly-available Redis codebase.
Furthermore, the number of buckets is constant and set to 16,384. Knowledge of these parameters places us in the setting of an S2 adversary.
Since a classic Redis application separates the server holding the sketch from the client interacting with the sketch, we consider that attack scenarios S3 or S4, while possible, are not typical for Redis. 
Thus, we only consider scenario S2 for Redis.
For simplicity, we set attack parameter $\epsilon$ to $\frac{1}{2}$ in the attack on an initially empty HLL sketch.
The results are averaged over 50 iterations and reported in Table~\ref{table:tab3}.
Similarly to the case for classic HLL, our empirical results match closely with our theoretical analysis.

\begin{table*}[tb!]
\begin{center}
\caption{Attack results for HLL in Redis, averaged over 50 iterations.}\label{table:tab3}
\begin{tabular}{| m{11em} | m{4em} | m{4.2em} |}
    \hline
    \textbf{Scenario} &  \multicolumn{2}{c|}{S2} \\ \hline
    %\textbf{HLL initially empty?} & no & yes \\ \hline
    \textbf{Original Card. Est.} & 0 & 139,967 \\ \hline
     \textbf{$\epsilon$} &  0.5 & - \\ \hline
    \textbf{\# Items added} & 524,210 & 524,454\\ \hline
    \textbf{Final Card. Est.} & 13,185 & 139,967 \\ \hline
\end{tabular}
\end{center}
\end{table*}
%\todo[inline]{Results from Redis or GBQ if we get them.}

@@

The authors thank Dennis Hofheinz and Anupama Unnikrishnan for useful discussions on simulation-based definitions, Tom Shrimpton for suggesting to look at GoogleBigQuery, Eylon Yogev for drawing our attention to related work on streaming algorithms with adversarial inputs, and Angelika Steger for help with balls-into-bins arguments.
\fi
